---
subtitle: "SMM under Partial Invariance"
---

```{r echo = FALSE}
## Thompson, M. & Green, S. (2013). Evaluating between-group differences
## in latent variable means. In G. Hancock & R. Mueller (Eds.), *Structural
## equation modeling: A second course* (2nd ed., pp. 163-218). Charlotte, NC:
## Information Age Publishing.
```

Thompson, M. & Green, S. (2013). Evaluating between-group differences in latent variable means. In G. Hancock & R. Mueller (Eds.), *Structural equation modeling: A second course* (2nd ed., pp. 163-218). Charlotte, NC: Information Age Publishing.

To assess differences in latent means under partial invariance, T&G use the Day-care versus Home_care example, but analyse a different data set. Summary data is presented in Table 5.2 (p. 183). 


#### Load lavaan package and get the data

```{r echo = c(-4,-1), eval = FALSE}
## Load package
library(lavaan)

## Get data from Table 5.2
# Group 1 - Day-care
cov1 <- c(
 154.54,
  44.75,  90.23,
  40.98,  22.77,  78.76,	
  41.35,   2.83,   7.92, 220.12,
  23.28,   9.12,   0.75,  61.46, 159.88,
  47.08,  22.88,  16.05, 125.08,  84.31, 332.26
) 

means1 <- c(49.14, 82.60, 104.95, 78.58, 54.95, 119.91)
n1 <- 250

# Group 2 - Home-care
cov2 <- c(
 124.93,
  52.19,  80.67,
  64.45,  42.85,  83.56,
  59.95,  33.10,  38.34, 290.65,
  32.32,  16.09,  18.29, 124.71, 169.17,
  87.15,  39.70,  51.82, 174.20, 108.39, 355.22
)

means2 <- c(55.01, 81.22, 97.01, 72.67, 46.57, 124.28)
n2 <- 150

# Get the variable names
names <- c("V1", "V2", "V3", "V4", "V5", "V6")

## Combine into lists
cov <- list("Day-care" = cov1, "Home-care" = cov2)
means <- list(means1, means2)
n <- list(n1, n2)

## Get the co/variance matrices
cov <- lapply(cov, lav_getcov, names = names)
```

In Table 5.3 (p. 184-187), T&G present the steps they follow to assess the differences in latent means.

- Step 1: Establish Configural Invariance - Equivalence of forms
- Step 2: Establish Scalar (or Weak) Invariance - Equivalence of loadings
- Step 3: Establish Metric (or Strong) Invariance - Equivalence of intercepts

At Steps 2 and 3, they can establish only partial invariance.

T&G use the marker-variable method of identification - one loading for each construct is fixed to one in all groups. I will use the reference-group method - constrain the variance of each construct to one. The parameter estimates for the two methods will differ, but they should give identical fit statistics. 

Why use the reference-group method? Later, when testing whether on not to free loadings constrained to equality, it is easier if pairs of loadings are not constrained to one. 

**Step 1** in Table 5.3 specifies a model with configural invariance - equivalence of forms. I follow the model specification except the latent variances are constrained to one in both groups, and all loadings are freely estimated. Using `std.lv = TRUE` in the `sem()` function takes care of all constraints. 

```{r echo = c(-2,-1), eval = FALSE}

## Step 1 - equivalence of forms - configural invariance
# Model statement
m1 <- "
   # Measurement Model
   Academic =~ V1 + V2 + V3
   Social =~ V4 + V5 + V6
"

# Fit the model to each group separately
m1_g1_fit <- sem(m1, sample.cov = cov[[1]], sample.nobs = n[[1]], 
   sample.mean = means[[1]], std.lv = TRUE)
summary(m1_g1_fit, remove.unused = FALSE)

m1_g2_fit <- sem(m1, sample.cov = cov[[2]], sample.nobs = n[[2]], 
   sample.mean = means[[2]], std.lv = TRUE)
summary(m1_g2_fit, remove.unused = FALSE)

# A function to extract fit measure from output
GetFit <- function(fit, ...) {
    fitMeasures(fit, ...)}
measures = c("chisq", "df", "pvalue", "cfi", "srmr", "rmsea")

# Get fit measures for each group
# Compare with Step 1 in Table 5.4
GetFit(m1_g1_fit, measures)
GetFit(m1_g2_fit, measures)

# Fit the two-group model
# Compare with Step 1 in Table 5.4
config_fit <- sem(m1, sample.cov = cov, sample.nobs = n, 
   sample.mean = means, std.lv = TRUE)
summary(config_fit, remove.unused = FALSE)
GetFit(config_fit, measures)
```


**Step 2** in Table 5.3 specifies a metric (or weak) invariance model. Use `group.equal = c("loadings")` in the `sem()` function to constrain the loadings to equality across the two groups. The code below follows the sequence of tests in Table 5.4: 

- The metric invariance model is fit and fit measures are obtained
- The `anova()` function compares the (nested) models - `config_fit` with `metric_fit`
- (A function to obtain differences in fit statistics)
- Differences in fit statistics are obtained
- The `lavTestScore()` function tests for releasing equality constraints (ie, constraints imposed on loadings).

Note that in the summary of the model fit, the latent variances in group 1 are constrained to one, but they are freely estimated in group 2. The latent means remain constrained to zero in both groups.

```{r echo = c(-2,-1), eval = FALSE}

## Step 2 - equivalence of loadings - metric (or weak) invariance

# Fit the model
# Compare with Step 2 in Table 5.4
metric_fit <- sem(m1, sample.cov = cov, sample.nobs = n, 
   sample.mean = means, std.lv = TRUE,
   group.equal = c("loadings"))
summary(metric_fit, remove.unused = FALSE)
GetFit(metric_fit, measures)

# Compare the fits of two models
anova(config_fit, metric_fit)

# Function to get difference in fit statistics
FitDiff = function(fit1, fit2) {
     call <- as.list(match.call())
     tab = cbind(data.frame(GetFit(fit1, measures), GetFit(fit2, measures)))[4:6, ]  
     names(tab) = c(call$fit1, call$fit2)
     tab$diff = tab[, 2] - tab[, 1]
     return(round(tab, 3))   
     }
FitDiff(config_fit, metric_fit)

# lavTestScore() function - test for releasing equality constraints
lavTestScore(metric_fit)

# Get parameter table
tab <- parTable(metric_fit)
# Print the part with loadings and constraints
tab[tab$op %in% c("=~","=="), ]
# Allow loadings for V3 to vary across groups
```


Note that `lavTestScore()` suggests that the equality constraint between `.p3.` and `.p26.` should be released first. The `parTable` helps to track down lavaan's labels: `.p3.` is the label for the loading of V3 on Academic in Group 1, and `.p26.` is the loading of V3 on Academic in Group 2. Thus, the loading of V3 on Academic is allowed to vary across the groups. 

In **Step 2a**, the equality constraint for the loadings of V3 on Academic are relaxed using `group.partial = c("Academic =~ V3")` in the `sem()` function. The steps are the same as before: the revised model is fit and fit statistics are obtained; the fit of the model is compared to the fit of the previous model; check for the next best equality constraint to relax.


```{r echo = c(-2,-1), eval = FALSE}

## Step 2a - relax equality constraint for V3 loading on Academic

# Fit the model 
# Compare with Step 2a in Table 5.4
metric_m2a_fit <- sem(m1, sample.cov = cov, sample.nobs = n, 
   sample.mean = means, std.lv = TRUE,
   group.equal = c("loadings"), group.partial = c("Academic =~ V3"))
summary(metric_m2a_fit, remove.unused = FALSE)
GetFit(metric_m2a_fit, measures)

anova(metric_fit, metric_m2a_fit)
FitDiff(metric_fit, metric_m2a_fit)

lavTestScore(metric_m2a_fit)
tab <- parTable(metric_m2a_fit)
tab[tab$op %in% c("=~","=="), ]
# Allow loadings for V6 to vary across groups
```

Relax the equality constraint between `.p6.` and `.p29.`; that is, in **Step 2b**, the equality constraint for the loadings of V6 on Social are relaexd.


```{r echo = c(-2,-1), eval = FALSE}

## Step 2b - relax equality constraint for V6 loading on Social

# Fit the model
# Compare with Step 2b in Table 5.4
metric_m2b_fit <- sem(m1, sample.cov = cov, sample.nobs = n, 
   sample.mean = means, std.lv = TRUE,
   group.equal = c("loadings"), group.partial = c("Academic =~ V3", "Social =~ V6"))
summary(metric_m2b_fit, remove.unused = FALSE)

GetFit(metric_m2b_fit, measures)
anova(metric_m2a_fit, metric_m2b_fit)
FitDiff(metric_m2a_fit, metric_m2b_fit)

lavTestScore(metric_m2b_fit)
tab <- parTable(metric_m2a_fit)
tab[tab$op %in% c("=~","=="), ]
# Allow loadings for V1 to vary across groups
```

Relax either the loadings for V1 or for V2 on Academic. T&G choose the constraint on V1. That is, in **Step 2c**, the equality constraints for V1 on Academic are relaxed. 

```{r echo = c(-2,-1), eval = FALSE}

## Step 2c - relax equality constraint for V1 loading on Social

# Fit the model
# Compare with Step 2c in Table 5.4
metric_m2c_fit <- sem(m1, sample.cov = cov, sample.nobs = n, 
   sample.mean = means, std.lv = TRUE,
   group.equal = c("loadings"), group.partial = c("Academic =~ V3", "Social =~ V6", "Academic =~ V1"))
summary(metric_m2c_fit, remove.unused = FALSE)

GetFit(metric_m2c_fit, measures)
anova(metric_m2b_fit, metric_m2c_fit)
FitDiff(metric_m2b_fit, metric_m2c_fit)
```

In the model carried forward into **Step 3**, the equality constraints for the loadings of V3 and V6 on their respective constructs are relaxed; and thus only partial scalar invariance has been established.

In **Step 3**, the intercepts are constrained to equality across the groups; that is, the argument for `group.equal` is `c("loadings", "intercepts")`, but with the equality constraints for loadings of V3 and V6 relaxed in the argument for `group.partial`. Because the constraints on the loading for V3 and V6 are relaxed, the equality constraints for the intercepts of V3 and V6 are also relaxed.

Note that in the summary of the model fit, the latent means are fixed to zero in Group 1, but are freely estimated in Group 2.

```{r echo = c(-2,-1), eval = FALSE}

## Step 3 -  equivalence of loadings & intercepts - scalar (strong) invariance

# Fit the model
# Compare with Step 3 in Table 5.4
scalar_m3_fit <- sem(m1, sample.cov = cov, sample.nobs = n, 
   sample.mean = means, std.lv = TRUE,
   group.equal = c("loadings", "intercepts"), 
   group.partial = c("Academic =~ V3", "Social =~ V6", "V3 ~ 1", "V6 ~ 1"))
summary(scalar_m3_fit, remove.unused = FALSE)
GetFit(scalar_m3_fit, measures)

anova(metric_m2b_fit, scalar_m3_fit)
FitDiff(metric_m2b_fit, scalar_m3_fit)

lavTestScore(scalar_m3_fit)
tab <- parTable(scalar_m3_fit)
tab[tab$op %in% c("=~","~1","=="), ]
```

Relaxing the intercept for V1 or V2 would improve the fit. T&G select V1. Thus in **Step 3a**, the equality constraint on the intercept for V1 is relaxed.

```{r echo = c(-2,-1), eval = FALSE}

## Step 3a - relax equality constraint for V1 intercept

# Fit the model
# Compare with Step 3a in Table 5.4
scalar_m3a_fit <- sem(m1, sample.cov = cov, sample.nobs = n, 
   sample.mean = means, std.lv = TRUE,
   group.equal = c("loadings", "intercepts"), 
   group.partial = c("Academic =~ V3", "Social =~ V6", "V3 ~ 1", "V6 ~ 1", "V1 ~ 1"))
summary(scalar_m3a_fit, remove.unused = FALSE)
GetFit(scalar_m3a_fit, measures)

anova(scalar_m3a_fit, scalar_m3_fit)
FitDiff(scalar_m3_fit, scalar_m3a_fit)

lavTestScore(scalar_m3a_fit)
tab <- parTable(scalar_m3a_fit)
tab[tab$op %in% c("=~","~1","=="), ]
```

Relaxing the intercept for V4 or V5 would improve the fit.  T&G select V4. Thus in **Step 3b**, the equality constraint on the intercept for V4 is relaxed.

```{r echo = c(-2,-1), eval = FALSE}

## Step 3b - relax equality constraint for V1 intercept

# Fit the model
# Compare with Step 3b in Table 5.4
scalar_m3b_fit <- sem(m1, sample.cov = cov, sample.nobs = n, 
   sample.mean = means, std.lv = TRUE,
   group.equal = c("loadings", "intercepts"), 
   group.partial = c("Academic =~ V3", "Social =~ V6", "V3 ~ 1", "V6 ~ 1", "V1 ~ 1", "V4 ~ 1"))
summary(scalar_m3b_fit, remove.unused = FALSE)
GetFit(scalar_m3b_fit, measures)

anova(scalar_m3a_fit, scalar_m3b_fit)
FitDiff(scalar_m3a_fit, scalar_m3b_fit)
```

The model carried forward into **Step 4** has equality constraints on V3 and V6 loadings relaxed, and equality contraints of intercepts for V1, V3, V4, and V6 relaxed.

In Step 4, attention turns to the latent means. In **Step 4a**, the means for the Academic construct are constrained to be equal, then in **Step 4b**, the means for Social are constrained to be equal. In each case, the fit is compared with the fit for the `scalar_m3b_fit`.


```{r echo = -1, eval = FALSE}

## Step 4a - constrain Academic mean to zero in Group 2
#  ie, Academic means constrained to equality

# Model statement
m4a <- "
  # Measurement Model
  Academic =~ V1 + V2 + V3
  Social =~ V4 + V5 + V6
  
  Academic ~ c(0,0)*1
  Social ~ c(0,NA)*1
"

scalar_m4a_fit <- sem(m4a, sample.cov = cov, sample.nobs = n, 
   sample.mean = means, std.lv = TRUE,
   group.equal = c("loadings", "intercepts"), 
   group.partial = c("Academic =~ V3", "Social =~ V6", "V3 ~ 1", "V6 ~ 1", "V1 ~ 1", "V4 ~ 1"))
summary(scalar_m4a_fit, remove.unused = FALSE)

GetFit(scalar_m4a_fit, measures)
anova(scalar_m3b_fit, scalar_m4a_fit)
FitDiff(scalar_m3b_fit, scalar_m4a_fit)


## Step 4b - constrain Social mean to zero in Group 2
#  ie, Social means constrained to equality

# Model statement
m4b <- "
  # Measurement Model
  Academic =~ V1 + V2 + V3
  Social =~ V4 + V5 + V6

  Academic ~ c(0,NA)*1
  Social ~ c(0,0)*1
"

scalar_m4b_fit <- sem(m4b, sample.cov = cov, sample.nobs = n, 
   sample.mean = means, std.lv = TRUE,
   group.equal = c("loadings", "intercepts"), 
   group.partial = c("Academic =~ V3", "Social =~ V6", "V3 ~ 1", "V6 ~ 1", "V1 ~ 1", "V4 ~ 1"))
summary(scalar_m4b_fit, remove.unused = FALSE)

GetFit(scalar_m4b_fit, measures)
anova(scalar_m3b_fit, scalar_m4b_fit)
FitDiff(scalar_m3b_fit, scalar_m4b_fit)
```

Conclusion: The means for both the Academic and the Social constructs differ across groups. 

How big are these differences? Focus turns to effect sizes. The formula for effect size are given on p. 195: difference in means divided by square root of variance. Again, T&G nominate the Day-care group as a reference group, and thus use its variance in the calculation of effect size. 


```{r echo = c(-4:-1), eval = FALSE}

## Get effect sizes (Day-care group as reference-group)
# Get means and variances from summary for scalar_m3b_fit
# Compare with effect sizes on p. 195
summary(scalar_m3b_fit)

ES_Acad <- (-0.266 - 0) / sqrt(1); ES_Acad
ES_Soc <- (-1.348 - 0) / sqrt(1); ES_Soc
```

Compare with effect sizes given on p. 195.

If the pooled variance is required, then, with unequal group sizes, pooled variance is a weighted average of the two variances.

If $n_1$ is the size of Group 1, and $n_2$ is the size of Group2, then the pooled variance is given by:
 
$$
 \sigma^2_p = \frac{(n_1 - 1)\sigma^2_1 + (n_2 - 1)\sigma^2_2}{n_1 + n_2 - 2}
$$

```{r eval = FALSE}
## Get effect sizes (pooled variance)
summary(scalar_m3b_fit)

ES_Acad <- (-0.266 - 0) / sqrt(((n1-1)*1 + (n2-1)*1.092)/(n1+n2-2)); ES_Acad
ES_Soc <- (-1.348 - 0) / sqrt(((n1-1)*1 + (n2-1)*1.986)/(n1+n2-2)); ES_Soc
```

As before, it might be better to get R to extract the means and variances from a list of estimates. Means are contained in element "alpha", and variances are contained in element "psi". 

```{r echo = -1, eval = FALSE}
## Extract means and variances from fitted lavaan object
#  Get estimates in a list
estimates <- lavInspect(scalar_m3b_fit, "est"); estimates

## Get effect sizes (pooled variances)
#  Get difference in means
means <- abs(Reduce('-', lapply(estimates, '[[', 'alpha'))); means

#  Get pooled variance
var <- lapply(lapply(estimates, '[[', 'psi'), diag); var
pool <- Reduce(function(x,y) ((n[[1]] - 1)*x + (n[[2]] - 1)*y) / (n[[1]] + n[[2]] - 2), var); pool

#  Get effect sizes
ES <- means/sqrt(pool)
colnames(ES) <- "ES"
ES
```

<br>


```{r}
#| echo: false
#| include: false
#| eval: true
#| purl: false
input <- knitr::current_input()
output <- "R/Thompson2_2013.r"
knitr::purl(input, output, documentation = 0, quiet = TRUE)
xfun::gsub_file(output, "^# NA", "")
xfun::gsub_file(output, "^# ", "")
```

```{r}
#| echo: true
#| code-fold: true 
#| code-summary: "R code with minimal commenting"
#| file: !expr "noquote(output)"
```