---
pagetitle: "Data from PDFs"

code-fold: true
---

:::: {layout="[60, 40]" layout-valign="center"}

::: {#leftcolumn}
<h2 style = "margin: 0px;"> Getting data out of PDF files into R</h2>
:::

::: {#rightcolumn}
  <img style = "margin: 0px; padding: 0px;" src="logo/logo.svg">
:::

:::: 


This post contains **R** scripts to extract data from PDFs of published papers and books. The aim is to get summary data (means, standard deviations, correlations, variances, covariances) from tables or table-like formats so that analyses can be reproduced. 

Two approaches are demonstrated. The first uses the `pdf_text()` function from the **pdftools** package. It is good at extracting text, but that means it will extract all text, whether contained in a table or not.  My approach is to get each line of text into a vector, then select the relevant lines. The data still needs tidying up - for that I use regular expressions. 

The second approach uses the `extract_tables()` function from the **tabulapdf** package. It is good at detecting and extracting a table. The extracted table usually needs some tidying up, for instance, to separate means and standard deviations from correlations, or to remove significance markers from correlations, but tidying-up is relatively straightforward.  When it works, it works well; but it does not always work. For instance, in Little et al., the data are not contained in a table, and not surprisingly, `extract_tables()` cannot find it. A more troubling example is shown in Thompson et al. The data are in a table but `extract_tables()` does not extract all rows of the table - not sure why. In these situations, **tabulapdf** provides an alternative function, `extract_text()`. As the name suggests, it extracts text - all text. The tidying-up process is then similar to that for **pdftools**.

The **tabulapdf** package requires **rJava** which implies a requirement for Java. It can be somewhat daunting having to install and set-up Java. The process can be simplified using the **rJavaEnv** package. Once installed, one line of **R** code will download, install, and setup Java for a given project.


#### The publications

::: {.no-indent}
- Kurbanoglu, N. & Takunyaci, M. (2021). A structural equation modeling on relationship between self-efficacy, physics laboratory anxiety and attitudes. *Journal of Family, Counseling and Education*, *6*(1), 47-56. </p>
The paper presents a basic three-variable mediation analysis. The R scripts extract means, standard deviations, and correlations for the three variables. The data are contained in Table 1 (p. 50). Arguably, it would be quicker and simpler to cut-and-paste the means, standard deviations, and correlations into R. The purpose is to outline the process for a simple case. 

```{r}
#| code-summary: "Kurbanoglu_pdftools.r"
#| file: "Kurbanoglu_2021/Kurbanoglu_pdftools.r"
```  

```{r}
#| code-summary: "Kurbanoglu_tabulapdf.r"
#| file: "Kurbanoglu_2021/Kurbanoglu_tabulapdf.r"
``` 
:::
 
<br>

::: {.no-indent}
- Little, T., Slegers, D., & Card, N. (2006). A non-arbitrary method of identifying and scaling latent variables in SEM and MACS models. *Structural Equation Modeling*, *13*(1), 59-72.  </p>
The paper presents a two-group, two-construct factor model. Each construct is assessed with three manifest indicators. The R scripts extract means, standard deviations, and correlations for each group, not from a table but from LISREL script (Appendix A, pp. 71-72). 
:::


```{r}
#| code-summary: "Little_pdftools.r"
#| file: "Little_2006/Little_pdftools.r"
```  

```{r}
#| code-summary: "Little_tabulapdf.r"
#| file: "Little_2006/Little_tabulapdf.r"
``` 



<br>

::: {.no-indent}
- Thompson, M. & Green, S. (2013). Evaluating between-group differences in latent variable means. In G. Hancock & R. Mueller (Eds.), *Structural equation modeling: A second course* (2nd ed., pp. 163-218). Charlotte, NC: Information Age Publishing.  </p>
The chapter presents two analyses assessing group differences in latent means using a two-group, two-construct factor model. Each construct is assessed with three manifest indicators. The data are contained in Table 5.1 (p. 169) and Table 5.2 (p. 183). The R scripts extract means and (co)variances for each group. There is also a MIMIC model with the data in the lower part of the relevant tables. I am not concerned with the MIMIC models, but I attempt to extract the data from Table 5.2 using **tabulapdf** only to show that **tabulapdf** struggles to extract the data cleanly.
:::

```{r}
#| code-summary: "Thompson_Table1.r"
#| file: "Thompson_2013/Thompson_Table1.r"
```  

```{r}
#| code-summary: "Thompson_Table2.r"
#| file: "Thompson_2013/Thompson_Table2.r"

```{r}
#| code-summary: "Thompson_Table1_Comb.r"
#| file: "Thompson_2013/Thompson_Table1_Comb.r"
```  


<br>

::: {.no-indent}
- Berry, D., & Willoughby, M. (2016). On the practical interpretability of cross-lagged panel models: Rethinking a developmental workhorse. Child Development, 88(4), 1186-1206. </p>
In Example 1 (commencing on page 14), the paper presents an Autoregressive Latent Trajectory Model with Structured Residuals. There are four waves of data for two variables.  The data are contained in the 2nd study in: </p>
Lansford, E., Criss, M., Laird, R., Shaw, D., Pettit, G., Bates, J., & Dodge, K. (2011). Reciprocal relations between parents' physical discipline and children's externalizing behavior during middle childhood and adolescence. *Development and Psychopathology*, *23*(1), 225–238.  </p>
The means and standard deviations are contained in Table 4 (p. 232), and the correlations are contained in Table 5 (p. 232).
:::


```{r}
#| code-summary: "Lansford_Study2.r"
#| file: "Berry_2016/Lansford_Study2.r"
```  
