---
title: "Methods of scaling applied to easier examples"
format: 
  html:
    css: styles.css
    code-fold: true
    embed-resources: true

execute:
  eval: false
  echo: true

highlight-style: solarized
---

<h4>A one-factor model.</h4>

<details>
  <summary>Model for Reference-Group Method</summary>
  <img src="../Drawing_SEMs/Little_2006/OneFactor1.svg">  
</details> 

<details>
  <summary>Model for Marker-Variable Method</summary>
  <img src="../Drawing_SEMs/Little_2006/OneFactor2.svg">  
</details> 

<details>
  <summary>Model for Effects-Scaling Method</summary>
  <img src="../Drawing_SEMs/Little_2006/OneFactor3.svg">  
</details> 

```{r}
#| code-summary: "OneFactor.r"
#| file: "Little_2006/OneFactor.r"
```  


<h4>A two-factor model</h4>

<details>
  <summary>Model for Reference-Group Method</summary>
  <img src="../Drawing_SEMs/Little_2006/TwoFactor1.svg">  
</details> 

<details>
  <summary>Model for Marker-Variable Method</summary>
  <img src="../Drawing_SEMs/Little_2006/TwoFactor2.svg">  
</details> 

<details>
  <summary>Model for Effects-Scaling Method</summary>
  <img src="../Drawing_SEMs//Little_2006/TwoFactor3.svg">  
</details> 

```{r}
#| code-summary: "TwoFactor.r"
#| file: "Little_2006/TwoFactor.r"
```  


<h4>A one-factor two-group model</h4>

<details>
  <summary>Model for Reference-Group Method</summary>
    <img src="../Drawing_SEMs/Little_2006/OneFactorTwoGroups1.svg">  
</details> 

<details>
  <summary>Model for Marker-Variable Method</summary>
    <img src="../Drawing_SEMs/Little_2006/OneFactorTwoGroups2.svg">  
</details> 

<details>
  <summary>Model for Effects-Scaling Method</summary>
    <img src="../Drawing_SEMs/Little_2006/OneFactorTwoGroups3.svg">  
</details> 

```{r}
#| code-summary: "OneFactorTwoGroup.r"
#| file: "Little_2006/OneFactorTwoGroup.r"
```  

