---
highlight-style: solarized
embed-resources: true
css: styles.css
code-fold: true

execute:
  eval: false
---

:::: {layout="[60, 40]" layout-valign="center"}

::: {#leftcolumn}
<h2 style = "margin: 0px;"> Reproducing published Structural Equation Models with OpenMx </h2>
:::

::: {#rightcolumn}
  <img style = "margin: 0px; padding: 0px;" src="logo/logo.svg">
:::

::::

This post presents **R** scripts to reproduce published Structural Equation Modeling analyses using **OpenMx**. 

Click on the **R** file to see the R script.
Click on the model to see a diagram of the model.

#### The publications

- Jose, P. (2013). *Doing statistical mediation and moderation*. New York, NY: Guilford Press. <br/> A basic three-variable mediation analysis (See Chapter 3).

<details class="image-fold">
  <summary>Mediation model</summary>
  <img src="../Drawing_SEMs/Jose_2013/Jose_2013.svg">  
</details> 

```{r}
#| code-summary: "Jose_2013.r"
#| file: "Jose_2013/Jose_2013.r"
```  

<br/>

- Kurbanoglu, N. & Takunyaci, M. (2021). A structural equation modeling on relationship between self-efficacy, physics laboratory anxiety and attitudes. *Journal of Family, Counseling and Education*, 6(1), 47-56. <br/> A basic three-variable mediation analysis using summary data.

<details class="image-fold">
  <summary>Mediation model</summary>
  <img src="../Drawing_SEMs/Kurbanoglu_2021/Kurbanoglu_2021.svg">  
</details> 

```{r}
#| code-summary: "Kurbanoglu_2021.r"
#| file: "Kurbanoglu_2021/Kurbanoglu_2021.r"
```  

<br/>
 
- Little, T., Slegers, D., & Card, N. (2006). A non-arbitrary method of identifying and scaling latent variables in SEM and MACS models. *Structural Equation Modeling*, *13*(1), 59-72. <br/> Three methods of scaling (reference group, marker variable, and effects scaling) in a two-factor, two-group model. Some easier examples (one-factor model, two-factor model, one-factor two-group model) are shown [here](EasierExamples.qmd).

<details class="image-fold">
  <summary>Model for Reference-Group Method</summary>
  <img src="../Drawing_SEMs/Little_2006/Scaling1.svg">  
</details> 

<details class="image-fold">
  <summary>Model for Marker-Variable Method</summary>
  <img src="../Drawing_SEMs/Little_2006/Scaling2.svg">  
</details> 

<details class="image-fold">
  <summary>Model for Effects-Scaling Method</summary>
  <img src="../Drawing_SEMs/Little_2006/scaling3.svg">  
</details> 

```{r}
#| code-summary: "Little_Scaling.r"
#| file: "Little_2006/Little_Scaling.r"
```  

<br/> 

- Thompson, M., Lie, Y. & Green, S. (2023). Flexible structural equation modeling approaches for analyzing means. In R. Hoyle (Ed.), *Handbook of structural equation modeling* (2nd ed., pp. 385-408). New York, NY: Guilford Press. <br/>
SEMs approaches for ANOVA and MANOVA type models.

<h5>The Models</h5>

<details>
  <summary>one-way ANOVA model</summary>
  <img src="../Drawing_SEMs/Green_2023/one_way_ANOVA.svg">  
</details> 

<details>
  <summary>one-way ANCOVA model</summary>
  <img src="../Drawing_SEMs/Green_2023/one_way_ANCOVA.svg">  
</details> 

<details>
  <summary>two-way ANOVA model</summary>
  <img src="../Drawing_SEMs/Green_2023/two_way_ANOVA.svg">  
</details> 

<details>
  <summary>one-way MANOVA model</summary>
  <img src="../Drawing_SEMs/Green_2023/one_way_MANOVA.svg">  
</details>
 
<details>
  <summary>one-way LATENT model</summary>
  <img src="../Drawing_SEMs/Green_2023/one_way_LATENT.svg">  
</details> 

<h5>The Data</h5>

```{r}
#| code-summary: "satisfactionI.r"
#| file: "Green_2023/satisfactionI.r"
```

```{r}
#| code-summary: "ANOVA_data.r"
#| file: "Green_2023/ANOVA_data.r"
```

```{r}
#| code-summary: "satisfactionII.r"
#| file: "Green_2023/satisfactionII.r"
```

<h5>R Scripts</h5>

```{r}
#| code-summary: "one_way_ANOVA.r"
#| file: "Green_2023/one_way_ANOVA.r"
```

```{r}
#| code-summary: "one_way_ANCOVA.r"
#| file: "Green_2023/one_way_ANCOVA.r"
```

```{r}
#| code-summary: "two_way_ANOVA.r"
#| file: "Green_2023/two_way_ANOVA.r"
```

```{r}
#| code-summary: "one_way_MANOVA.r"
#| file: "Green_2023/one_way_MANOVA.r"
```

```{r}
#| code-summary: "one_way_LATENT.r"
#| file: "Green_2023/one_way_LATENT.r"
```
